---
title: "Reproducible Research Assignment 1"
author: "Alana Souter"
date: "26 August 2017"
output: html_document
---

# Reproducible Research Assignment 1

## Prepare Environment

```{r chunk0}
library(ggplot2)
library(knitr)
opts_chunk$set(echo=TRUE, results='hold')
```

## Loading and Preprocessing the Data

1. Load the Data

```{r chunk1}
Data <- read.csv('activity.csv', header=TRUE, sep=",",colClasses=c("numeric","character","numeric"))
```

2. Tidying the Data

```{r chunk2}
Data$date<-as.Date(Data$date,format="%Y-%m-%d")
Data$interval<-as.factor(Data$interval)
str(Data)
```

## What is the mean total number of steps taken per day?

```{r chunk3}
StepsPerDay<-aggregate(steps~date,Data,sum)
colnames(StepsPerDay)<-c("date","steps")
head(StepsPerDay)
```

1. Make a histogram of the total number of steps taken each day

```{r histogram1}
ggplot(StepsPerDay, aes(x=steps))+
  geom_histogram(fill="blue",binwidth=1000)+
  labs(title="Steps Taken per Day", x="Number of Steps per Day", y= "Number of times in day")+ theme_bw()
```

2. Calculate the mean and median total number of steps taken per day

```{r chunk4}
MeanSteps<-mean(StepsPerDay$steps, na.rm=TRUE)
MedianSteps<-median(StepsPerDay$steps, na.rm=TRUE)
```

## What is the average daily activity pattern?
```{r chunk5}
AvgStepsPerInterval<-aggregate(Data$steps,
                               by=list(interval=Data$interval),
                               FUN=mean, na.rm=TRUE)
AvgStepsPerInterval$interval<-as.integer(levels(AvgStepsPerInterval$interval)[AvgStepsPerInterval$interval])
colnames(AvgStepsPerInterval)<-c("interval","steps")
```

1. Make a time series plot

```{r plot1}
ggplot(AvgStepsPerInterval,aes(x=interval,y=steps))+
  geom_line(color="red",size=1)+
  labs(title="Average Daily Activity Pattern", x="Interval",y="Number of Steps")+
  theme_bw()
```

2. Which 5 minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r chunk6}
MaximumInterval<-AvgStepsPerInterval[which.max(AvgStepsPerInterval$steps),]
```

## Inputting missing values

1. Calculate the total number of missing values in the dataset

```{r chunk7}
MissingValues<-sum(is.na(Data$steps))
```

2. Devise a strategy for filling in all of the missing values in the dataset & 3. Create new dataset that is equal to the original dataset but with the missing data filled in.

```{r chunk8}
NAfill <- function(data, pervalue) {
        na_index <- which(is.na(data$steps))
        na_replace <- unlist(lapply(na_index, FUN=function(idx){
                interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        fill_steps <- Data$steps
        fill_steps[na_index] <- na_replace
        fill_steps
}

Datafill <- data.frame(  
        steps = NAfill(Data, AvgStepsPerInterval),  
        date = Data$date,  
        interval = Data$interval)
str(Datafill)
```

4. Make a histogram of the total number of steps taken each day

```{r histogram2}
StepsPerDayFilled<-aggregate(steps~date,Datafill,sum)
colnames(StepsPerDayFilled)<-c("date","steps")
ggplot(StepsPerDayFilled, aes(x=steps))+
  geom_histogram(fill="blue",binwidth=1000)+
  labs(title="Steps Taken per Day", x="Number of Steps per Day", y= "Number of times in a Day")+theme_bw()
```

Calculate and report the mean and median total number of steps taken per day.

```{r chunk9}
StepsPerDayFillMean<-mean(StepsPerDayFilled$steps,na.rm=TRUE)
StepsPerDayFillMedian<-median(StepsPerDayFilled$steps,na.rm=TRUE)
```

## Are there differences in activity patterns between weekdays and weekends?

```{r chunk10}
weekdays_steps <- function(data) {
    weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
                          FUN=mean, na.rm=T)
    # convert to integers for plotting
    weekdays_steps$interval <- 
            as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
    colnames(weekdays_steps) <- c("interval", "steps")
    weekdays_steps
}

data_by_weekdays <- function(data) {
    data$weekday <- 
            as.factor(weekdays(data$date)) # weekdays
    weekend_data <- subset(data, weekday %in% c("Saturday","Sunday"))
    weekday_data <- subset(data, !weekday %in% c("Saturday","Sunday"))

    weekend_steps <- weekdays_steps(weekend_data)
    weekday_steps <- weekdays_steps(weekday_data)

    weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
    weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))

    data_by_weekdays <- rbind(weekend_steps, weekday_steps)
    data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
    data_by_weekdays
}

data_weekdays <- data_by_weekdays(Datafill)
```

2. Make a panel plot containing a time series plot

```{r plot2}
ggplot(data_weekdays, aes(x=interval, y=steps)) + 
        geom_line(color="red") + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of steps") +
        theme_bw()
```


